---
title: 'Software Internals: Science-Wise False Discovery Rate'
author: "Nathan (Nat) Goodman"
date: "June 1, 2017"
output:
  html_document:
    code_folding: hide
    css: css/html_document.css
    highlight: kate
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: 3
linkcolor: cyan
citecolor: green
urlcolor: blue
---
```{r echo=FALSE,include=FALSE,cache=FALSE}
## setup variables for documentation paths
knitr::read_chunk('../tool/doclink.R');
knitr::read_chunk('../tool/codelink.R');
```
```{r echo=FALSE,include=FALSE,cache=FALSE}
<<doclink>>
<<codelink>>
```

*This document describes the software design and design choices of the scripts in [my GitHub SWFDR repository](https://github.com/natgoodman/SWFDR). The base script is `swfdr_base.R`, a simple implementation that uses base R capabilities only. Other scripts extend the base implementation by providing solutions to some of the [Exercises for the Reader](`r exercises`). 
Please see the companion document [Science: Science-Wise False Discovery Rate](`r science`) for a discussion of the conceptual background.* 

**Presently just copy of software.Rmd.bak2. Will be cut back to "Software" section of that document.**

## Introduction

The *false discovery rate* (*FDR*) is the probability that a significant p-value indicates a false positive, or equivalently, the proportion of significant p-values that correspond to results without a real effect. FDR estimation is routinely used in genomics for multiple test correction in large scale experiments. 

Some time ago I wrote a quick-and-dirty R script that compared FDR estimations methods used in genomics on simulated data. I decided to adapt this program to explore the notion of science-wise false discovery rate as explained in the companion document [Science: Science-Wise False Discovery Rate](`r science`). 

Beyond the interesting science, I thought this would be a good example to get my feet wet in the world of didactic R software aimed at web users. I cleaned up the code, splitting off a simple core and casting the rest as [Exercises for the Reader](`r exercises`). I put the code in a [GitHub repository](https://github.com/natgoodman/SWFDR) and wrote R style and more conventional documentation. Along the way, I expanded my paltry knowledge of GitHib mechanics and learned R documentation tools from scratch. 

I hope the software will be a good example for readers who have some R experience but aren’t experts. The base script, `swfdr_base.R`, is a simple implementation; the others extend the base in small steps. Most of the scripts use base R capabilities only and will run "out of the box" on any reasonably modern R installation. The software (mostly) conforms to good software engineering practice: it uses a simple, consistent coding style with lots of comments, and consists of small functions with well-defined purposes and clear interfaces (i.e., inputs and outputs). 

The distribution provides the software as a collection of functions meant for interactive use, similar to what a non-expert R user is likely to write. It also provides the software as R packages which can be loaded as source packages (via `devtools::load_all`) or installed and loaded by `library` the usual way. The main advantage of the package form is that the R help system will work, e.g., you can type `help(run)` to get help on the `run` function.

I encourage reader-programmers to download the code, try it out, and modify it to do more. To this end, please see [Exercises for the Reader](`r exercises`) for ideas on possible extensions. The software is open source, released under the MIT license, which means you can do almost anything you want with it. I’d love to see your improvements if you choose to share them, but you are under no obligation to do so.

## Where's the code

- The `R/` subdirectory of the distribution contains the raw source code in script form
- The `pkg/` subdirectory contains the code in package form
- For each script, there are also HTML and PDF document containing the code as formatted text. E.g. for `swfdr_base.R`, the code documents are [doc/swfdr_base/code.html](swfdr_base/code.html) and [doc/swfdr_base/code.pdf](swfdr_base/code.pdf).

## Scenario

The program simulates a large number of problem instances representing published results, some of which are true and some false. The instances are very simple: I generate two groups of random numbers and use the t-test to assess the difference between their means. One group (the control group or simply *group0*) comes from a standard normal distribution with $mean=0$. The other group (the treatment group or simply *group1*) is a little more involved:

* for *true* instances, we take numbers from a standard normal distribution with mean *d* ($d>0$);
* for *false* instances, we use the same distribution as *group0*.

The parameter *d* is the effect size, aka *Cohen’s d*. 

I use the t-test to compare the means of the groups and produce a p-value assessing whether both groups comes from the same distribution. 

The program does this thousands of times (drawing different random numbers each time, of course), collects the resulting p-values, and computes the FDR. The program repeats the procedure for a range of assumptions to determine the conditions under which most published results are wrong.

For *true* instances, we expect the difference in means to be approximately *d* and for *false* ones to be approximately 0, but due to the vagaries of random sampling, this may not be so. If the actual difference in means is far from the expected value, the t-test may get it wrong, declaring a *false* instance to be positive and a *true* one to be negative. The goal is to see how often we get the wrong answer across a range of assumptions.

## Nomenclature

To reduce confusion, I will be obsessively consistent in my terminology.

* An *instance* is a single run of the simulation procedure.
* The terms *positive* and *negative* refer to the results of the t-test. A *positive instance* is one for which the t-test reports a significant p-value; a *negative instance* is the opposite. Obviously the distinction between positive and negative rests on the chosen significance level.
- *true* and *false* refer to the correct answers. A *true instance* is one where the treatment group (*group1*) is drawn from a distribution with $mean=d$ ($d>0$). A *false instance* is the opposite: an instance drawn from a distribution with $mean=0$.
* *empirical* refers to results calculated from the simulated data, as opposed to *theoretical* which means results calculated using standard formulas.

## Programming Style

As you examine the software, you'll no doubt notice that my R code looks different from other examples out there and eschews many recommendations in well-respected R style guides by [Hadley Wickham](http://adv-r.had.co.nz/Style.html) and [Google](https://google.github.io/styleguide/Rguide.xml). One obvious difference is that my code is scrunched with less whitespace than others. In addition, I use `=` instead of `<-`, prefer single-quotes over double, end statements with semi-colons, start full line comments with ##, and use `.` as a word separator in variable names and `_` in function names. I have good reasons for some of these choices, but mostly it’s just habit.

Programming style is a matter of taste (unless your boss foists a style on you). My style works for me. As you gain experience, I encourage you to find a style that works for you.

## Software 

``` {r include=FALSE}
## set global scriptname for codelink
scriptname='swfdr_base'
knitr::read_chunk(paste(sep='','../tool/','chunk.',scriptname,'.R'));
```

The plan is to go through the base script, `swfdr_base.R`, in detail to see a complete working example, then look at select bits of the others.

First, a logisitical digression. 

The HTML version of this document presents code examples in "folded" code blocks as well as links that take you to the right place in the code document. 
`r if (out=='html') paste(sep='',"Here's an example. To reveal the code, click the \"Code\" button at the left hand margin below or [follow this link](swfdr_base/code.html","). (In the example, the link takes you to the top of the code document but in real cases it will take you directly to the code.)")`

``` {r include=out=='html',eval=FALSE}
## Example code block
## Now that you see it, you can re-hide the code by clicking the "Hide" button above
```

The PDF version is less code-friendly. PDF does not support code folding, so I decided to omit long code blocks that would disrupt the flow. Nor does PDF support links to points inside a document, so I can't provide links that take you to the right place in the code document. A reasonable practice, assuming your PDF viewer allows it, is to leave the code document open, display the table of contents in the sidebar, and navigate to code sections as needed. 

### `swfdr_base.R`


#### `run`
The top-level function is `run`. The code block below shows how to invoke `run` with default parameters assuming your working directory is the root of the repository. `r codelink(link=F)`
``` {r eval=FALSE,include=!nocode}
source("script/swfdr_base.R");
run();
```

`run` has 4 steps. `r codelink('run')`
``` {r eval=FALSE,include=!nocode}
<<run>>
```

* `init` sets everything up
* `doit` (pronounced “do it!”) does all the work
* `saveit` (pronounced “save it!”) optionally saves the parameters, figures, and results
* `fini` does final cleanup if any; there’s nothing to do in this example, but I include it for stylistic consistency.

#### Global Variables
A key consideration is this sort of program is how to pass parameters and results from one step to the next. I decided to use the very simple scheme of passing everything in global variables. This runs counter to conventional wisdom in the programming world but is okay if it's well-documented. It’s also conducive for interactive use.

The table below lists the global variables. The first group are simulation parameters, next are parameters that control program operation, and last are results.

| variable | meaning | default or source |
| ------------- | -------------------------------------------- | ------------- |
| prop.true | fraction of cases where there is a real effect | `seq(.1,.9,by=.2)` |
| m | number of iterations | `1e4` |
| n | sample size | `16` |
| d | standardized effect size (aka *Cohen's d*) | `c(.25,.50,.75,1,2)` |
| pwr | power. if set, program adjusts *d* to achieve power | `NA` |
| sig.level | significance level for power calculations when *pwr* is set | `0.05` |
| pval.plot | p-values for which we plot results | `c(.001,.01,.03,.05,.1)` |
|  |  | |
| scriptname | used to set output directories and in error messages |`'swfdr_base'` |
| datadir | data directory relative to distribution root | `'data/swfdr_base'` |
| figdir | figure directory relative to distribution root | `'figure/swfdr_base'` |
| save | save parameters, results, and plots; <br/>sets *save.rdata* and *save.plot*, not *save.txt* | `FALSE` |
| save.rdata | save parameters and results in RData format | `FALSE` (set by *save*)|
| save.txt | save results in txt format. **CAUTION: big and slow** | `FALSE` (not set by *save*) |
| save.plot | save plots | `FALSE` (set by *save*)|
| clean | remove contents of data and figure directories; <br/>sets *clean.data* and *clean.fig* | `FALSE` |
| clean.data | remove contents of data directory | `FALSE` (set by *clean*) |
| clean.fig | remove contents of figure directory | `FALSE` (set by *clean*) |
|  |  | |
| cases | parameter grid | created by `init` |
| sim | simulation results | created by `dosim` |
| interp | interpolation results | created by `dointerp ` |

#### Initialize and Finalize
`init` sets the parameters. `r codelink('init',nocode=TRUE)`
``` {r eval=FALSE,include=!nocode}
<<init>>
```

`init`'s argument list defines all parameters and provides default values. To change a parameter, simply call `run` (or `init` directly) with the parameter and new value, e.g., `run(n=32)` changes the sample size *n* from its default (16) to the new value (32). `init` creates a parameter grid, called *cases*, containing all combinations of parameters. The code section at the bottom of `init` copies all parameters and its result, cases, into global variables.

For the default parameters, the parameter grid expands to 25 cases (5 values of *prop.true* $\times$ 5 values of *d*; all other parameters have single-valued defaults). The program does 10,000 simulations for each case for a total of 250,000 simulations. This takes about 3 minutes on my small Linux server.

`fini` does final cleanup if any. There’s nothing to do in this example, but I include it for stylistic consistency. `r codelink('fini',nocode=TRUE)`
``` {r eval=FALSE,include=!nocode}
<<fini>>
```

#### `doit`
`doit` has 3 steps. `r codelink('init')`
``` {r eval=FALSE,include=!nocode}
<<doit>>
```

* `dosim` runs the simulation
* `dointerp` interpolates relevant columns of the simulation results at fixed p-values so we can plot results for p-values of interest
* `doplot` plots the results and optionally save the plots.

#### Simulation Functions
`dosim` considers cases one by one, calling `sim_one` to do the heavy lifting, and stores the result in a global data frame, called *sim*. `r codelink('dosim')`
``` {r eval=FALSE,include=!nocode}
<<dosim>>
```

`sim_one` does the actual simulations. `r codelink('sim_one',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<sim_one>>
```

The program draws *m* random samples of size *n* for each of the two groups, storing the numbers for each group in a $m \times n$ matrix whose rows are instances and columns are samples. The samples for *group0* are from a standard normal distribution with $mean=0$, i.e. `rnorm(n,mean=0)`. For *group1*, the code computes the requisite numbers of true and false instances (*num.true* and *num.false*, resp.), draws *num.true* samples from a standard normal distribution with $mean=d$, i.e. `rnorm(n,mean=d)` and *num.false* samples from the same distribution as *group0*.

The program runs the t-test on each of the *m* pairs of samples, generating *m* p-values, and calculates theoretical and empirical FDRs for each p-value. It also calculates and stores basic statistics for each pair of samples, e.g., the actual means and differences, should this be of interest later.

`sim_one` calls functions `fdr_theo` and `fdr_empi` to do the FDR calculations. 

#### FDR Calculations
Recall that FDR is the number of false positives divided by the total number of positives, and of course, the total number of positives is the number of false positives plus the number of true ones. In pseudo-math: $FDR = num.fp / (num.fp + num.tp)$. 

The theoretical FDR uses textbook formulas to estimate *num.fp* and *num.tp* for each p-value *pval*. `r codelink('fdr_theo')`
``` {r eval=FALSE,include=!nocode}
<<fdr_theo>>
```

- *num.fp* = *pval* times the number of false instances = $pval \times (1-prop.true) \times m$
- *num.tp* = power (given *pval*) times the number of true instances = $power \times prop.true \times m$

The program calculates power using R’s power.t.test function: `power.t.test(n=n,delta=d,sd=1,sig.level=p)$power`.

The empirical FDR is conceptually simple, but the code is complicated by the possibility of duplicate p-values. `r codelink('fdr_empi')`
``` {r eval=FALSE,include=!nocode}
<<fdr_empi>>
```

The simple part is the code before and after the inner `if`: for each p-value *pval*, the code counts the number of false positives in instances with p-values $\leq pval$ and divides by the number of all instances with such p-values. To make this calculation easy, the function first sorts by p-value, computes the answer, then reorders the result back to the original order. The complexity comes in the inner `if`: in the unlikely, but possible, event that multiple instances have the same p-value, the FDR calculation would probably give different answers for each duplicate. The code handles this case by setting the empirical FDR to the maximum value for all instances with the given p-value. 

#### Interpolation Functions
The structure of `dointerp` is similar to `dosim`. It calls `interp_one` on each case to interpolate the theoretical and empirical FDRs at desired p-values using R’s `approxfun` function. `dointerp` stores the result in a global data frame, called *interp*. `r codelink('dointerp')`
``` {r eval=FALSE,include=!nocode}
<<dointerp>>
<<interp_one>>
```

#### Plot Functions
These functions operate on the `sim` and `interp` data frames produced by `dosim` and `dointerp`
respectively.

`doplot` is the main plot function. It calls separate functions for each of the four kinds of plot. `r codelink('doplot')`
``` {r eval=FALSE,include=!nocode}
<<doplot>>
```

- `plot_byprop` plots FDR by *prop.true* for one value of *d*. `r codelink('plot_byprop',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<plot_byprop>>
```

- `plot_byd` plots FDR by *d* for one value of *prop.true*. `r codelink('plot_byd',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<plot_byd>>
```

- `plot_vsprop` plots FDR vs *prop.true* for one value of *d* at fixed p-values. `r codelink('plot_vsprop',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<plot_vsprop>>
```

- `plot_vsd` plots FDR vs *d* for one value of *prop.true* at fixed p-values. `r codelink('plot_vsd',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<plot_vsd>>
```

The four plot functions are similar. Each function plots a different slice of theoretical and empirical FDR as a function of three variables: $FDR=f(prop.true,d,pval)$. The plot functions differ in how they reduce four dimensions (FDR and the three variables) to something that can be plotted in two dimensions.  

Each function starts by fixing one variable to a single value. Next, the function splits the data into groups based on a second variable. Finally, it plots each group vs. the remaining variable, using different line types (solid vs dashed) to distinguish theoretical and empirical FDR. 

Much of the code in each function deals with plotting mechanics: setting up the axes, title, legend, and grid lines, and optionally saving the plots. 

`plot_byprop` and `plot_byd` operate on *sim*; `plot_vsprop` and `plot_vsd` operate on *interp*.

#### Save and Load
`saveit` saves all parameters and results (actually, all global variables) in RData and optionally tab-delimited text formats. `r codelink('saveit',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<saveit>>
```

The companion function, `loadit`, loads the saved data into the R session. `r codelink('loadit',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<loadit>>
```
#### Utility Functions
There are two utility (helper) functions. `r codelink('power_grid',nocode=T)`
``` {r eval=FALSE,include=!nocode}
<<power_grid>>
<<colramp>>
```
- `power_grid` adds power calculations to the *cases* data frame
- `colramp` sets up the colors for the plot functions.

### more scripts TBD

## Conclusion or Wrap-up TBD ##
