```{r echo=FALSE, eval=FALSE}
#################################################################################
## R markdown User Guide child document generated from 
## script/swfdr_base.Rxl and script/swfdr_base.R by tool/rdoc.pl.
#################################################################################
##
## Author:  Nat Goodman
## Copyright (C) 2017 Nat Goodman
##
## This software is open source, distributed under the MIT License. See LICENSE
## file at https://github.com/natgoodman/NewPro/FDR/LICENSE
##
#################################################################################
```


### `init` - *Initialize program parameters.*

#### Description

Processes parameters and stores them in global variables. Creates parameter grid, called `cases`, containing all combinations of  parameters. Creates output directories if they do not exist.

#### Usage

```{r eval=FALSE}
init(prop.true = seq(0.1, 0.9, by = 0.2),
     m = 10000,
     n = 16,
     d = c(0.25, 0.5, 0.75, 1, 2),
     pwr = NA, sig.level = 0.05,
     pval.plot = c(0.001, 0.01, 0.03, 0.05, 0.1),
     scriptname = "swfdr_base",
     datadir = file.path("data", scriptname),
     figdir = file.path("figure", scriptname),
     save = F, save.rdata = save, save.txt = F, save.plot = save,
     clean = F, clean.data = clean, clean.fig = clean)
```

#### Arguments
```{r echo=FALSE, eval=FALSE}
| argument      | meaning                                      | default       |
| ------------- | -------------------------------------------- | ------------- |
| `prop.true` | fraction of cases where where there is a real effect. | `seq(0.1, 0.9, by = 0.2)` |
| `m` | number of iterations. | `1e4` |
| `n` | sample size. | `16` |
| `d` | standardized effect size (aka *Cohen's d*) | `c(0.25, 0.5, 0.75, 1, 2)` |
| `pwr` | power. if set, program adjusts `d` to achieve power. | `NA` |
| `sig.level` | significance level for power calculation | `0.05` |
| `pval.plot` | p-values for which we plot results | `c(1e-03, 0.01, 0.03, 0.05, 0.1)` |
| `scriptname` | script name. Used to construct output directory path names. | `"swfdr_base"` |
| `datadir` | path name of directory for data files. | `"data/swfdr_base"` |
| `figdir` | path name of directory for plots. | `"figure/swfdr_base"` |
| `save` | logical. sets `save.rdata` and `save.plot` | `FALSE` |
| `save.rdata` | if TRUE, save parameters and results (actually, all global variables) in RData format. The output filename is `globals.RData` in directory `datadir`. | `FALSE` (set by `save`) |
| `save.txt` | save simulation and interpolation results as tab-delimited text files. The output filenames are `sim.txt` amd `interp.txt` in directory `datadir`. **CAUTION: big & slow!** | `FALSE` |
| `save.plot` | if TRUE, save plots as png files in `figdir`. The output filenames are `plot_byd.png`, `plot_byprop.png`, `plot_vsd.png`, `plot_vsprop.png` | `FALSE` (set by `save`) |
| `clean` | logical. sets `clean.data` and `clean.fig`. | `FALSE` |
| `clean.data` | if TRUE, delete contents of datadir and start fresh | `FALSE` (set by `clean`) |
| `clean.fig` | if TRUE, delete contents of figdir and start fresh | `FALSE` (set by `clean`) |

```
```{r child=argtable('init.swfdr_base.argtable')}
```

#### Value

The `cases` data frame is returned invisibly.

#### Details

For the default parameters, the `cases` parameter grid expands to 25 cases (5 values of prop.true x 5 values of d; all other parameters have single-valued defaults). We do 10,000 simulations for each case for a total of 250,000 simulations. This takes about 3 minutes on my small Linux server.

This function is usually called by `run`. It may be called directly if the user wishes to perform custom initialization.

#### Examples

```{r eval=FALSE}
# initialize parameters with default values
init();

# initialize parameters with default values but save results in directories data/guide01 and plots in figure/guide01
init(save=T,datadir='data/guide01',figdir='figure/guide01');

# initialize parameters with values requiring less runtime by reducing number of simulation runs and simulated cases
init(m=1e3,d=c(0.25,0.5,1),prop.true=c(0.3,0.5,0.8));

# specify power directly and let program adjust effect size
init(m=1e3,pwr=c(0.1,0.3,0.8),prop.true=c(0.3,0.5,0.8));
```
