```{r echo=FALSE, eval=FALSE}
#################################################################################
## R markdown User Guide child document generated from 
## script/swfdr_base.Rxl and script/swfdr_base.R by tool/rdoc.pl.
#################################################################################
##
## Author:  Nat Goodman
## Copyright (C) 2017 Nat Goodman
##
## This software is open source, distributed under the MIT License. See LICENSE
## file at https://github.com/natgoodman/NewPro/FDR/LICENSE
##
#################################################################################
```
<span id=doplot></span><span id=plot_byprop></span><span id=plot_byd></span><span id=plot_vsprop></span><span id=plot_vsd></span>

### Plot Functions - *Plot the results*

#### Description

These functions operate on the `sim` and `interp` data frames produced by `dosim` and `dointerp` 
respectively.

#### Usage

```{r eval=FALSE}
doplot=function(save.plot=F)

plot_byprop=function(save.plot=F,d1=1,sig.level=.05)

plot_byd=function(save.plot=F,prop.true1=0.5,sig.level=.05)

plot_vsprop=function(save.plot=F,d1=1,sig.level=.05)

plot_vsd=function(save.plot=F,prop.true1=0.5,sig.level=.05)
```

#### Arguments
```{r echo=FALSE, eval=FALSE}
| argument      | meaning                                      | default       |
| ------------- | -------------------------------------------- | ------------- |
| `save.plot` | if TRUE, save the plot. The output format is PNG. The output filename is the name of the function with `.png` suffix  in directory `figdir`, eg, `figure/plot_byprop.png`. | `FALSE` |
| `d1` | fixed value of `d` for dimension reduction. If `d1` is not in the `d` vector, it is set d1 to `max(d)`. | `1` |
| `prop.true1` | fixed value of `prop.true` for dimension reduction. If `prop.true` is not in the `prop.true` vector, it is set to the first value in `prop.true`. | `0.5` |
| `sig.level` | values of p-value and FDR marked by dashed lines on the plot. | `0.05` |

```
```{r child=argtable('plot-functions.swfdr_base.argtable')}
```

#### Details

`doplot` is the main plot function. It calls separate functions for each of the four kinds of plot. 
* `plot_byprop` plots FDR by prop.true for one value of d
* `plot_byd` plots FDR by d for one value of prop.true
* `plot_vsprop` plots FDR vs prop.true for one value of d at fixed p-values
* `plot_vsd` plots FDR vs d for one value of prop.true at fixed p-values

Each of the `plot_` functions plots a different slice of theoretical and empirical FDR as a function of three variables: FDR=f(prop.true,d,pval) The functions differ in how they reduce four dimensions (FDR and the three variables) to something that can be plotted in two dimensions.

Each function starts by fixing one variable to a single value. Next, the function splits the data into groups based on a second variable. Finally, it plots each group vs. the remaining variable, using different line types (solid vs dashed) to distinguish theoretical and empirical FDR.

`plot_byprop` and `plot_byd` operate on `sim`; `plot_vsprop` and `plot_vsd` operate on `interp`.
